---
title: "Mouse E18 brain data analysis"
author:
- name: Michal Kabza
  affiliation:
    - Roche Informatics, F. Hoffmann-La Roche Ltd., Poznań, Poland
  email: michal.kabza@contractors.roche.com
- name: Tim Sterne-Weiler
  affiliation:
    - Department of Oncology Bioinformatics, Genentech Inc., South San Francisco, USA
    - Department of Discovery Oncology, Genentech Inc., South San Francisco, USA
  email: sternewt@gene.com
output:
  BiocStyle::html_document:
    toc_depth: 1
    toc_float: true
    df_print: paged
    code_folding: show
package: Isosceles
vignette: >
  %\VignetteIndexEntry{Mouse E18 brain data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

# Introduction 

This vignette demonstrates how Isosceles can be used to identify isoforms and 
PSI events showing differences in expression along or between different cell 
lineages or trajectories. The analysis is based on mouse E18 brain scRNA-Seq 
data from the Lebrigand et al. 2020 paper - all the details regarding how the 
data was processed can be found in the `mouse_E18_analysis` directory of the 
[Isosceles_Paper](https://github.com/Genentech/Isosceles_Paper) repository.

Load the required packages:
```{r, message = FALSE}
library(Isosceles)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(scuttle)
library(DEXSeq)
library(RColorBrewer)
library(pheatmap)
library(dittoSeq)
library(Nebulosa)
```

# Data exploration

Load the Isosceles TCC SummarizedExperiment object:
```{r, message = FALSE, warning = FALSE}
se_tcc <- readRDS(system.file(
    "extdata", "se_tcc_mouse_e18.rds",
    package = "Isosceles"
))
```

Load the processed SingleCellExperiment objects (containing normalized data, UMAP coordinates etc.):
```{r, message = FALSE, warning = FALSE}
sce_gene <- readRDS(system.file(
    "extdata", "sce_gene_mouse_e18.rds",
    package = "Isosceles"
))
sce_transcript <- readRDS(system.file(
    "extdata", "sce_transcript_mouse_e18.rds",
    package = "Isosceles"
))
sce_psi <- readRDS(system.file(
    "extdata", "sce_psi_mouse_e18.rds",
    package = "Isosceles"
))
```

Load the matrix pf pseudotime values for individual trajectories calculated using the Slingshot package:
```{r, message = FALSE, warning = FALSE}
pseudotime_matrix <- readRDS(system.file(
    "extdata", "pseudotime_matrix_mouse_e18.rds", package = "Isosceles"
))
head(pseudotime_matrix)
```

This dataset contains the following cell types and trajectories, shown on the UMAP plot below:

  * Glutamatergic neuron lineage with two trajectories starting from the same progenitor cells:
    * Trajectory 1 (glut_1, clusters glut.4, glut.8, glut.5 and glut.7) 
    * Trajectory 2 (glut_2, clusters glut.4, glut.11 and glut.9)
  * GABAergic neuron trajectory (gaba, clusters gaba.2 and gaba.3)
  * Radial glia (rad_glia, cluster rad_glia.6)
  * Cycling radial glia (cyc_rad_glia, cluster cyc_rad_glia.1)
  * Cajal-Retzius cells (cr, cluster cr.10)

```{r, message = FALSE, warning = FALSE, class.source = 'fold-hide', fig.dim = c(7, 5)}
cluster_colors <- c(
    glut.4 = brewer.pal(n = 7, name = "YlGnBu")[3],
    glut.8 = brewer.pal(n = 7, name = "YlGnBu")[4],
    glut.5 = brewer.pal(n = 7, name = "YlGnBu")[5],
    glut.7 = brewer.pal(n = 7, name = "YlGnBu")[6],
    glut.11 = brewer.pal(n = 7, name = "YlGn")[5],
    glut.9 = brewer.pal(n = 7, name = "YlGn")[6],
    gaba.2 = brewer.pal(n = 7, name = "Reds")[4],
    gaba.3 = brewer.pal(n = 7, name = "Reds")[6],
    rad_glia.6 = colorRampPalette(c("white", "deeppink"))(7)[6],
    cyc_rad_glia.1 = colorRampPalette(c("white", "purple"))(7)[6],
    cr.10 = colorRampPalette(c("white", "gold"))(7)[6]
)
umap_df <- as.data.frame(reducedDim(sce_gene, "UMAP"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")
cluster_umap_df <- umap_df %>%
    mutate(cluster = sce_gene$cluster) %>%
    group_by(cluster) %>%
    summarise(
        UMAP_1 = mean(UMAP_1),
        UMAP_2 = mean(UMAP_2)
    )
umap_lineages <- list(
    glut_1 = c("glut.4", "glut.8", "glut.5", "glut.7"),
    glut_2 = c("glut.4", "glut.11", "glut.9"),
    gaba = c("gaba.2", "gaba.3")
)
lineage_umap_df <- umap_lineages %>%
    enframe() %>%
    unchop(value) %>%
    dplyr::rename(lineage = name, cluster = value) %>%
    left_join(cluster_umap_df)
set.seed(103)
dittoDimPlot(sce_gene, "cluster",
             reduction.use = "UMAP", main = "",
             size = 2, do.label = TRUE, labels.size = 4,
             labels.highlight = TRUE, legend.show = FALSE) +
    scale_colour_manual(values = cluster_colors) +
    geom_path(
        data = lineage_umap_df,
        size = 0.8,
        mapping = aes(x = UMAP_1, y = UMAP_2, group = lineage),
        arrow = arrow(ends = "last", type = "closed",
                      length = unit(0.1, "inches"))
    )
```

Cluster cell type identities have been established using the marker gene 
expression signatures from the Lebrigand et al. 2020 paper:
```{r, message = FALSE, warning = FALSE, class.source = 'fold-hide', fig.dim = c(10, 7)}
marker_sets <- list(
    Progenitors = c("Neurog2", "Eomes"), # intermediate progenitors
    Glut = c("Neurod6", "Neurod2"), # glutamatergic
    Mat_Glut = c("Camk2b", "Opcml", "Crym"), # mature glutamatergic
    GABA = c("Gad2", "Gad1", "Dlx2"), # GABAergic
    Mat_GABA = c("Maf", "Mafb", "Arx"), # mature GABAergic
    Rad_glia = c("Fabp7", "Vim", "Dbi"), #radial glia
    Cyc_rad_glia = c("Cenpf", "Top2a", "Mki67"), # cycling radial glia
    CR_cells = c("Snhg11", "Lhx5", "Reln") # Cajal-Retzius cells
)
sce_gene_marker <- sce_gene
rownames(sce_gene_marker) <- make.unique(
    rowData(sce_gene_marker)$gene_name, sep = "@"
)
cell_marker_scores <- sumCountsAcrossFeatures(
    sce_gene_marker, marker_sets,
    exprs_values = "logcounts", average = TRUE
)
sce_gene_set <- SingleCellExperiment(
    assays = list(logcounts = cell_marker_scores),
    colData =  colData(sce_gene_marker)
)
reducedDim(sce_gene_set, "UMAP") <- reducedDim(sce_gene_marker, "UMAP")
plot_density(
    sce_gene_set,
    c("Progenitors", "Glut", "Mat_Glut", "GABA", "Mat_GABA",
      "Rad_glia", "Cyc_rad_glia", "CR_cells"),
    slot = "logcounts", size = 1
)
```

# Isoform switching

Isosceles allows for easy detection of isoform switching events by comparing 
every pair of cell groups (e.g. clusters) and identifying transcripts of the 
same gene showing statistically significant differences in opposite directions:
```{r, message = FALSE, warning = FALSE}
isoswitch_df <- find_isoswitch(
    se = sce_transcript,
    cell_labels = sce_transcript$cluster,
    min_fdr = 0.01
)
```

Print the isoform switching results:
```{r, message = FALSE, warning = FALSE}
isoswitch_df
```

Plot the expression density the Clta gene and its transcripts involved in isoform switching:
```{r, message = FALSE, warning = FALSE, class.source = 'fold-hide', fig.dim = c(8, 5)}
p1 <- plot_density(
        sce_gene, "ENSMUSG00000028478",
        size = 1.5
    ) +
    labs(title = "Clta (ENSMUSG00000028478)") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
p2 <- plot_density(
        sce_transcript,
        "ISOT-6683-8d1f-209c-002d:s44012650:e44032850:AP:FL",
        size = 1.5
    ) +
    scale_color_gradientn(colours = brewer.pal(n = 7, name = "Reds")) +
    labs(title = "Clta-204 (ENSMUST00000107849)") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
p3 <- plot_density(
        sce_transcript,
        "ISOT-eeca-a52b-651e-905d:s44012600:e44032850:AP:FL",
        size = 1.5
    ) +
    scale_color_gradientn(colours = brewer.pal(n = 7, name = "Blues")) +
    labs(title = "Clta-206 (ENSMUST00000170241)") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
p4 <- plot_density(
        sce_transcript,
        "ISOT-8a93-5996-c87a-8c5f:s44004450:e44032850:AP:FL",
        size = 1.5
    ) +
    scale_color_gradientn(colours = brewer.pal(n = 7, name = "Greens")) +
    labs(title = "Clta-202 (ENSMUST00000107846)") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
(p1 + p2) / (p3 + p4)
```

# DEXSeq analysis

Isosceles can also be used to aggregate PSI event counts at the pseudotime 
window level and create a DEXSeqDataSet object suitable for the analysis of 
their changes along a pseudotime trajectory (the workflow can be easily modified 
to allow for any method of cell aggregation or experiment design). Given the 
number of PSI events  found in annotations, we recommend filtering them 
(e.g. by the number of cells showing signs of the event's inclusion) - in this 
vignette we'll restrict the analysis to the row names of the (already 
pre-filtered) `sce_psi` object. We can prepare a DEXSeqDataSet object for the
glutamatergic trajectory 1 using the following commands:
```{r, message = FALSE, warning = FALSE}
# Aggregate TCC values using moving window over pseudotime
se_window_tcc <- pseudotime_tcc(
    se_tcc = se_tcc,
    pseudotime = pseudotime_matrix[, "glut_1"],
    window_size = 30,
    window_step = 15
)
se_window_gene <- tcc_to_gene(
    se_tcc = se_window_tcc
)
se_window_transcript <- tcc_to_transcript(
    se_tcc = se_window_tcc,
    use_length_normalization = FALSE
)
se_window_psi <- transcript_to_psi(
    se = se_window_transcript,
)
se_window_psi <- add_psi_counts(
    se_psi = se_window_psi,
    se_gene = se_window_gene
)
# Prepare the DEXSeqDataSet object
window_pseudotime <- as.numeric(scale(se_window_psi$pseudotime))
dxd <- psi_to_dexseq(
    se_psi = se_window_psi,
    condition = window_pseudotime,
    psi_events = rownames(sce_psi)
)
```

Print the DEXSeqDataSet object:
```{r, message = FALSE, warning = FALSE}
dxd
```

Standard DEXSeq workflow can be used for further analysis:
```{r, message = FALSE, warning = FALSE}
dxd <- estimateSizeFactors(dxd)
dxd <- estimateDispersions(dxd)
dxd <- testForDEU(dxd)
dxd <- estimateExonFoldChanges(dxd, fitExpToVar = "condition")
```

Process the DEXSeq output to make it more readable:
```{r, message = FALSE, warning = FALSE}
# Process the results of the DEXSeqResults function
dexseq_results_df <- DEXSeqResults(dxd) %>%
    as.data.frame() %>%
    dplyr::select(featureID, pvalue, padj) %>%
    transmute(
        psi_event = featureID,
        gene_id = sapply(strsplit(featureID, ":"), "[", 1),
        gene_name = rowData(sce_gene[gene_id,])$gene_name,
        pvalue = pvalue,
        fdr = padj
    )
rownames(dexseq_results_df) <- NULL
# Calculate the maximum absolute value of logFC across 
# different pseudotime points
logFC_values <- as.data.frame(DEXSeqResults(dxd))
logFC_values <- logFC_values[, grepl("^log2fold_", colnames(logFC_values))]
logFC_values <- as.matrix(logFC_values)
max_abs_logFC <- apply(logFC_values, 1, function(x) {max(abs(x))})
max_abs_logFC[is.na(max_abs_logFC)] <- 0
names(max_abs_logFC) <- NULL
dexseq_results_df$max_abs_logFC <- max_abs_logFC
# Filter the DEXSeq results
dexseq_results_df <- filter(
    dexseq_results_df, fdr <= 0.05, max_abs_logFC >= 1
)
```

Print the DEXSeq results:
```{r, message = FALSE, warning = FALSE}
dexseq_results_df
```

Plot the expression density the Celf2 gene and the density of PSI values for 
its three selected PSI events significantly changing in the glutamatergic 
trajectory 1:
```{r, message = FALSE, warning = FALSE, class.source = 'fold-hide', fig.dim = c(8, 5)}
p1 <- plot_density(
        sce_gene, "ENSMUSG00000002107",
        size = 1.5
    ) +
    labs(title = "Celf2 (ENSMUSG00000002107)") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
p2 <- plot_density(
        sce_psi,
        "ENSMUSG00000002107:chr2:6560659-6560670:-:A5",
        slot = "psi",
        size = 1.5
    ) +
    scale_color_gradientn(colours = brewer.pal(n = 7, name = "Reds")) +
    labs(title = "chr2:6560659-6560670:-:A5") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
p3 <- plot_density(
        sce_psi,
        "ENSMUSG00000002107:chr2:6553965-6553982:-:A3",
        slot = "psi",
        size = 1.5
    ) +
    scale_color_gradientn(colours = brewer.pal(n = 7, name = "Blues")) +
    labs(title = "chr2:6553965-6553982:-:A3") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
p4 <- plot_density(
        sce_psi,
        "ENSMUSG00000002107:chr2:6546780-6547041:-:RI",
        slot = "psi",
        size = 1.5
    ) +
    scale_color_gradientn(colours = brewer.pal(n = 7, name = "Greens")) +
    labs(title = "chr2:6546780-6547041:-:RI") +
    theme(legend.position = "right",
          plot.title = element_text(size = 12),
          legend.title = element_text(size = 11))
(p1 + p2) / (p3 + p4)
```

In order to visualize the detected PSI events as a heatmap, Isosceles can
calculate the PSI count to mean  permuted PSI count ratios for pseudotime 
windows with smaller, trajectory-dependent window and step sizes (**warning**: 
the process will take ~30 min):
```{r, message = FALSE, warning = FALSE, eval = FALSE}
set.seed(100)
psi_ratio_matrix <- calculate_psi_ratio_matrix(
    se_tcc = se_tcc,
    pseudotime_matrix = pseudotime_matrix,
    psi_events = dexseq_results_df$psi_event,
    window_sizes = c(
        glut_1 = 100, glut_2 = 100, gaba = 100,
        rad_glia = 50, cyc_rad_glia = 50, cr = 50
    ),
    window_steps = c(
        glut_1 = 3, glut_2 = 3, gaba = 3,
        rad_glia = 3, cyc_rad_glia = 3, cr = 3
    ),
    n_perm = 10
)
```

```{r, include = FALSE}
psi_ratio_matrix <- readRDS("psi_ratio_matrix.rds")
```

Plot the PSI count ratio heatmap (trajectories are denoted using the same 
colors as in the trajectory UMAP plot in the *Data exploration* section): 
```{r, message = FALSE, warning = FALSE, class.source = 'fold-hide', fig.dim = c(7, 3.5)}
# Prepare the heatmap values
heatmap_values <- log2(psi_ratio_matrix + 0.1)
heatmap_values[is.nan(heatmap_values)] <- 0
# Prepare heatmap annotations
heatmap_trajectory <- sapply(
    strsplit(colnames(heatmap_values), "\\."), "[", 1
)
heatmap_gap_positions <- cumsum(rle(heatmap_trajectory)$lengths)
col_ann_df <- data.frame(
    trajectory = heatmap_trajectory
)
rownames(col_ann_df) <- colnames(heatmap_values)
col_ann_colors <- list(
    trajectory = c(
        glut_1 = brewer.pal(n = 7, name = "YlGnBu")[5],
        glut_2 = brewer.pal(n = 7, name = "YlGn")[5],
        gaba = brewer.pal(n = 7, name = "Reds")[5],
        rad_glia = colorRampPalette(c("white", "deeppink"))(7)[6],
        cyc_rad_glia = colorRampPalette(c("white", "purple"))(7)[6],
        cr = colorRampPalette(c("white", "gold"))(7)[6]
    )
)
# Replace PSI events IDs with more readable labels 
gene_id_to_name <- setNames(
    rowData(sce_gene)$gene_name, rowData(sce_gene)$gene_id
)
psi_labels <- sapply(
    strsplit(rownames(heatmap_values), "\\:"), function(x) {
        paste0(c(gene_id_to_name[x[1]], x[3], x[5]),
               collapse = ":")
    }
)
rownames(heatmap_values) <- psi_labels
# Plot the heatmap
pheatmap(
    heatmap_values,
    breaks = seq(-3, 3, length.out = 101),
    cluster_rows = TRUE, cluster_cols = FALSE,
    show_rownames = TRUE, show_colnames = FALSE,
    annotation_col = col_ann_df,
    annotation_colors = col_ann_colors,
    gaps_col = heatmap_gap_positions,
    legend = TRUE, annotation_legend = FALSE,
    annotation_names_col = FALSE,
    fontsize_row = 5, treeheight_row = 0,
    scale = "row"
)
```

# Session Info

```{r}
sessionInfo()
```
